Classical Simulation of Quantum Error Correction in a Fibonacci Anyon Code

### 2015-06-11 fib2.tex
some more minor rewording

We consider a 
2-dimensional system with
point-like anyonic quasiparticle excitations, which exhibit exotic statistics under exchange.
The worldlines of these exchanges form \emph{braids}
in (2+1)-dimensions
which act unitarily on (the Hilbert space of) degrees of freedom known as the \emph{fusion space.}
To access this space we define
the topological observables of the system: these
are total charge measurements within regions bounded by \dude{directed} closed loops (Wilson loops).
The possible results of such a charge measurement correspond to different anyon particle types in the model.
%The algebraic properties of these processes
These dynamics and observables obey algebraic rules
given by a unitary modular tensor category~\cite{Wang2010b}. % (UMTC). 
However, from the perspective of
\emph{topological quantum field theory}
anyons are modelled
as punctures in a surface~\cite{Pfeifer2014}, and it is not
the anyons themselves that braid,
but the manifold in which the anyons reside 
that becomes deformed (twisted) by these braid moves.
This is the perspective we adopt (see \Fref{f:braidloop}).
\dude{Fig or FIG ?}
%In particular, this shows the action of braiding
%on arbitrary observables (Wilson loops) as
%these deform with the manifold.

@@

%The state \dude{space?} of a collection of anyons
%can be described by the observables of the system.
The \dude{key?} difference between abelian and nonabelian
anyon theories is that outcomes of total charge measurements in an abelian theory
are determined uniquely by particle content only.
In contrast, outcomes for nonabelian charge measurements depend on the history of the particles as well as their type.
%The defining characteristic of a \emph{nonabelian}
%anyon theory is that ...
That is, the fusion space is one-dimensional for abelian anyons, and generally larger for nonabelian anyons.
In particular, we consider a system supporting (nonabelian) Fibonacci
anyon excitations, denoted by $\tau$.
Two such anyons can have total charge
that is either $\tau$ or $\vac$ (vacuum), or any superposition of these, and
so the fusion space in this case is 2-dimensional.
We can represent basis states for this space using diagrams 
of definite total charge for the Wilson loops, 
and arbitrary states as linear combinations of these diagrams:
\begin{align*}
\includegraphics[]{pic-fibonacci.pdf}
\end{align*}
For $n$ anyons of type $\tau$, the 
dimension of this space grows asymptotically as $\varphi^n$ for $\varphi=\frac{1+\sqrt{5}}{2}$ 
the golden ratio.

Observables associated to non-intersecting loops commute, and so
a basis for the space can be built from a maximal
set of disjoint, nested loops.
For three anyons, two
possible ways of nesting loops are related
via $F$-moves:
\begin{align}
\label{e:fmove}
\includegraphics[]{pic-fmove.pdf}
\end{align}
Here we have shown particles with charges $a$, $b$, and $c$
as well as total charge $d$.
The lefthand side shows a state where the anyons with charge $a$ and $b$ are
observed to have total charge $i$.
The righthand side involves a superposition 
over total charge $j$ of the anyons with charge $b$ and $c$.

The green line serves to linearly
order the anyons, and track
deformations in the underlying space 
(the history of braiding processes).
It also denotes the direction along
which $F$-moves occur.

A clockwise braiding process is represented by an $R$-move:
\begin{align}
\label{e:rmove}
\includegraphics[]{pic-rmove.pdf}
\end{align}
This is meant to show a \emph{half-twist} of the space
inside the observable with value $c$
holding two anyons (punctures) with charges $a$ and $b$. \cggb{This statement isn't quite true. I suggest just deleting.}

@@

For the Fibonacci anyons, the non-trivial $R$ and $F$ moves are 
\begin{equation*}
	R_{\vac}^{\tau\tau} = \e^{\frac{-4\pi i}{5}} 
	\ \ , \ \
	R_\tau^{\tau\tau}= \e^{\frac{3\pi i}{5}} 
	\ \ , \ \
	F_{\tau}^{\tau\tau\tau} = \begin{pmatrix}\varphi^{-1}&\varphi^{-\frac{1}{2}}\\\varphi^{-\frac{1}{2}}&-\varphi^{-1}\end{pmatrix} \,,
\end{equation*}
where the matrix is given in a basis labelled $(\vac,\tau)$. For more details of the Fibonacci anyon theory, see e.g.~Ref.~\cite{Nayak2008} and references therein. 
Braiding of Fibonacci anyons and measurements of fusion outcomes is known to be 
sufficient to implement universal quantum computation~\cite{Freedman2002, Nayak2008}.

Finally, in addition to the fusion space of any anyons present, there is also a global degeneracy associated with the topology of the 
manifold on which the anyons reside. We will consider systems with the topology of a torus, 
which for the Fibonacci anyons gives rise to a 2-fold degeneracy.
This global space is the one in which we will encode, and demonstrate protection of, quantum 
information.

@@

%We model the system on a periodic $L\times L$ square lattice $\Lambda$ of \emph{tiles} on which anyons can reside. 
%This treatment applies equally well to lattice models, where anyons 
%can only be located at a discrete set of points, 
%or continuum models, where a discretization of the manifold must 
%be chosen in order to perform charge measurements. 
%In the former case, each tile of $\Lambda$ represents a 
%lattice site, while in the latter case each tile represents 
%a measurement region. For more details of an analogous model, see~\cite{Brell2013}.

We consider encoding a qubit of quantum information in the global degeneracy associated 
with the topology of the manifold of our system.
We endow this manifold with a finite set of $L\times L$ Wilson loops arranged in
a square lattice tiling $\Lambda$.
%We tile the manifols with  a finite set of $L\times L$ Wilson loops arranged in
These \emph{tiles} will be the observables accesible to the error-correction procedure (\emph{decoder}).
We can use these observables to construct an idealized Hamiltonian for this system of the form
\begin{align}
	H=-\sum_{t\in \Lambda}\proj{\vac}_t\;,\label{e:hamiltonian}
\end{align}
with $\proj{\vac}_t$ the projector to charge $\vac$ at tile $t$, i.e.~the ground 
space of the model has vacuum total charge in each tile.
Given the toroidal boundary 
conditions, this space \cggb{is} the two-fold degenerate codespace. \cggb{Subtlety in that there could be anyons around in the ground space.}
%We identify this ground space as the codespace of our model.

@@

%In order to keep track of the state of the system we maintain a set of disjoint curves, 
%\cggb{do they need to be directed, or is that just for the figure?} 
%representing the history of anyons on the lattice.
%These curves can also be thought of as a top-down view of the fusion tree:
%the anyons comprising the fusion tree are here displayed as points along each curve. 
%In general, the curves wind around our two 
%dimensional manifold in a haphazard way determined by the progress of the simulation.
%This creates a dynamically generated basis for the fusion space 
%that easily allows for tensor factorization of disjoint components. 
%For comparison, previous work used a single fixed basis for 
%the fusion space determined by row-major ordering of sites~\cite{Brell2013}.

>>>>

We consider a 
2-dimensional system with
point-like anyonic quasiparticle excitations, which exhibit exotic statistics under exchange.
The worldlines of these exchanges form \emph{braids}
in (2+1)-dimensions
which act unitarily on the Hilbert space of degrees of freedom known as the \emph{fusion space}.
To access this space we define
the topological observables of the system: these
are total charge measurements within regions bounded by %\dude{directed} 
closed loops (Wilson loops).
The possible results of such a charge measurement correspond to different anyon particle types in the model.
%The algebraic properties of these processes
These dynamics and observables obey algebraic rules
given by a unitary modular tensor category~\cite{Wang2010b}. % (UMTC). 
However, from the perspective of
\emph{topological quantum field theory}
anyons are modeled
as punctures in a surface~\cite{Pfeifer2014}, and it is not
the anyons themselves that braid,
but the manifold in which the anyons reside 
that becomes deformed (twisted) by these braid moves.
This is the perspective we adopt (see \Fref{f:braidloop}).
%\dude{Fig or FIG ?}
%In particular, this shows the action of braiding
%on arbitrary observables (Wilson loops) as
%these deform with the manifold.

@@

%The state \dude{space?} of a collection of anyons
%can be described by the observables of the system.
The defining %\dude{key?} 
difference between abelian and nonabelian
anyon theories is that in an abelian theory particle content alone uniquely determines the outcome of total charge measurements.
In contrast, outcomes for nonabelian charge measurements depend on the history of the particles as well as their type.
%The defining characteristic of a \emph{nonabelian}
%anyon theory is that ...
That is, the fusion space is one-dimensional for abelian anyons, and generally larger for nonabelian anyons.
In particular, we consider a system supporting (nonabelian) Fibonacci
anyon excitations, denoted by $\tau$.
Two such anyons can have total charge
that is either $\tau$ or $\vac$ (vacuum), or any superposition of these, and
so the fusion space in this case is 2-dimensional.
We can represent basis states for this space using diagrams 
of definite total charge for the Wilson loops, 
and arbitrary states as linear combinations of these diagrams:
\begin{align*}
\includegraphics[]{pic-fibonacci.pdf}
\end{align*}
For $n$ anyons of type $\tau$, the 
dimension of this space grows asymptotically as $\varphi^n$, where $\varphi=\frac{1+\sqrt{5}}{2}$ 
is the golden ratio.

Observables associated to non-intersecting loops commute, and so
a basis for the space can be built from a maximal
set of disjoint, nested loops.
For three anyons, two
possible ways of nesting loops are related
via $F$-moves:
\begin{align}
\label{e:fmove}
\includegraphics[]{pic-fmove.pdf}
\end{align}
Here we have shown particles with charges $a$, $b$, and $c$
as well as total charge $d$.
The lefthand side shows a state where the anyons with charge $a$ and $b$ are
observed to have total charge $i$.
The righthand side involves a superposition 
over total charge $j$ of the anyons with charge $b$ and $c$. 
The green line serves to linearly
order the anyons, and track
deformations in the underlying space, 
i.e.\ the history of braiding processes.
It also denotes the direction along
which $F$-moves occur. A clockwise braiding process is represented by an $R$-move:
\begin{align}
\label{e:rmove}
\includegraphics[]{pic-rmove.pdf}
\end{align}
%This is meant to show a \emph{half-twist} of the space
%inside the observable with value $c$
%holding two anyons (punctures) with charges $a$ and $b$. \cggb{This statement isn't quite true. I suggest just deleting.}

@@

For Fibonacci anyons, the non-trivial $R$ and $F$ moves are 
\begin{equation*}
	R_{\vac}^{\tau\tau} = \e^{\frac{-4\pi i}{5}} 
	\ \ , \ \
	R_\tau^{\tau\tau}= \e^{\frac{3\pi i}{5}} 
	\ \ , \ \
	F_{\tau}^{\tau\tau\tau} = \begin{pmatrix}\varphi^{-1}&\varphi^{-\frac{1}{2}}\\\varphi^{-\frac{1}{2}}&-\varphi^{-1}\end{pmatrix} \,,
\end{equation*}
where the matrix is given in a basis labelled $(\vac,\tau)$. For more details of the Fibonacci anyon theory, see e.g.\ Ref.~\cite{Nayak2008} and references therein. 
%Braiding of Fibonacci anyons and measurements of fusion outcomes is known to be 
%sufficient to implement universal quantum computation~\cite{Freedman2002, Nayak2008}.
% -- we already said this in the intro and we're pressed for space, so I cut this sentence

In addition to the fusion space there is also a global degeneracy associated with the topology of the 
manifold on which the anyons reside. We consider systems with the topology of a torus, 
which for the Fibonacci anyons gives rise to a 2-fold degeneracy.
%This global space is the one in which we will encode, and demonstrate protection of, quantum information. 
% -- we say this again in the paragraph after next.

@@

%We model the system on a periodic $L\times L$ square lattice $\Lambda$ of \emph{tiles} on which anyons can reside. 
%This treatment applies equally well to lattice models, where anyons 
%can only be located at a discrete set of points, 
%or continuum models, where a discretization of the manifold must 
%be chosen in order to perform charge measurements. 
%In the former case, each tile of $\Lambda$ represents a 
%lattice site, while in the latter case each tile represents 
%a measurement region. 
%For more details of an analogous model, see~\cite{Brell2013}.
\cggb{I would consider restoring the sentence that refers the reader to our previous paper for details of the phenomenological model.}

We consider encoding a qubit of quantum information in the global degeneracy associated 
with the topology of the manifold of our system.
We endow this manifold with a finite set of $L\times L$ Wilson loops arranged in
a square lattice tiling $\Lambda$.
%We tile the manifold with a finite set of $L\times L$ Wilson loops arranged in
These \emph{tiles} will be the observables accesible to the error-correction procedure (\emph{decoder}).
We can use these observables to construct an idealized Hamiltonian for this system of the form
\begin{align}
	H=-\sum_{t\in \Lambda}\proj{\vac}_t\;,\label{e:hamiltonian}
\end{align}
with $\proj{\vac}_t$ the projector to charge $\vac$ at tile $t$, i.e.~the ground 
space of the model has vacuum total charge in each tile.
Given the toroidal boundary 
conditions, this space is the two-fold degenerate codespace. \cggb{Subtlety in that there could be anyons around in the ground space.}
%We identify this ground space as the codespace of our model.

@@

%In order to keep track of the state of the system we maintain a set of disjoint curves, 
%\cggb{do they need to be directed, or is that just for the figure?} 
%representing the history of anyons on the lattice.
%These curves can also be thought of as a top-down view of the fusion tree:
%the anyons comprising the fusion tree are here displayed as points along each curve. 
%In general, the curves wind around our two 
%dimensional manifold in a haphazard way determined by the progress of the simulation.
%This creates a dynamically generated basis for the fusion space 
%that easily allows for tensor factorization of disjoint components. 
%For comparison, previous work used a single fixed basis for 
%the fusion space determined by row-major ordering of sites~\cite{Brell2013}.
\cggb{I would be an advocate for restoring the sentences that compare the dynamically generated basis here to our previous work's fixed fusion basis.}
